---
title: "Cyclistic Bike-Share Analysis Case Study"
author: "Adiya Dangwal"
date: "2024-09-16"
output: html_document
---

## Introduction

Welcome to the Cyclistic bike-share analysis case study! In this case study, I am working as a junior data analyst at Cyclistic, a bike-share company based in Chicago. My role involves analyzing bike trip data to support the development of a new marketing strategy aimed at converting casual riders into annual members.

## Scenario

As a junior data analyst on the marketing analytics team at Cyclistic, my task is to understand the usage patterns of casual riders versus annual members. The director of marketing, Lily Moreno, believes that increasing the number of annual memberships is crucial for the company's future success. My analysis will help inform marketing strategies that could convert casual riders into annual members.

### Company Background

Cyclistic operates a bike-share program with more than 5,800 bicycles and 600 docking stations across Chicago. The program includes a variety of bikes, such as traditional, reclining, hand tricycles, and cargo bikes, making it inclusive for people with disabilities and riders who need alternative options. About 8% of riders use assistive bikes, while the majority use traditional bikes. Cyclistic users mainly ride for leisure, but around 30% use the bikes for commuting.

### Marketing and Business Goals

Cyclistic's marketing strategy has previously focused on general awareness and appealing to broad consumer segments through various pricing plans: single-ride passes, full-day passes, and annual memberships. Casual riders are those who purchase single-ride or full-day passes, whereas annual members are those with yearly subscriptions. Financial analysts have found that annual members are significantly more profitable than casual riders.

Lily Moreno aims to design a marketing strategy to convert casual riders into annual members. She believes that casual riders, who are already familiar with Cyclistic, are prime candidates for membership. To achieve this goal, we need to analyze historical bike trip data to identify key trends and insights.

## Data Analysis Process

To address the business questions and support the development of the marketing strategy, I will follow these steps of the data analysis process:

1. **Ask**: Define the key questions and objectives for the analysis.
2. **Prepare**: Gather and preprocess the data, including cleaning and structuring it for analysis.
3. **Process**: Conduct exploratory data analysis (EDA) to identify patterns and trends.
4. **Analyze**: Perform detailed analysis to compare the behavior of casual riders and annual members.
5. **Share**: Create visualizations and reports to present the findings.
6. **Act**: Formulate recommendations based on the insights and prepare a presentation for the executive team.

## Next Steps

I will start by preparing the data for analysis. This involves cleaning the data, exploring key variables, and setting up the environment for detailed analysis. I will also formulate specific questions to guide the analysis, such as:

- How do the usage patterns of casual riders differ from those of annual members?
- What factors contribute to the decision to purchase an annual membership?
- How can digital media strategies be optimized to target casual riders effectively?

By following this structured approach, I will ensure that the insights derived from the data are actionable and aligned with Cyclistic's marketing goals.

## Ask

### Business Task

The primary business task is to determine how annual members and casual riders use Cyclistic bikes differently. This understanding will help in crafting targeted marketing strategies to convert casual riders into annual members by highlighting the unique benefits of annual memberships and addressing the needs or preferences that drive usage patterns.

### Key Questions

1. **How do annual members and casual riders use Cyclistic bikes differently?**
   - To answer this question, we need to analyze usage patterns, trip frequencies, durations, and bike types preferred by each group.

2. **Why would casual riders buy Cyclistic annual memberships?**
   - Understanding the motivators and potential barriers for casual riders will guide marketing strategies to address these factors.

3. **How can Cyclistic use digital media to influence casual riders to become members?**
   - Identifying effective digital media strategies will help in targeting casual riders and promoting the benefits of annual memberships.

### Key Tasks

1. **Identify the Business Task**:
   - The task is to compare the bike usage patterns between annual members and casual riders. This comparison will provide insights into how different user groups engage with the bike-share program, which can inform marketing strategies aimed at converting casual riders to annual members.

2. **Consider Key Stakeholders**:
   - **Lily Moreno** (Director of Marketing): Interested in understanding user behavior to design effective marketing campaigns.
   - **Cyclistic Marketing Analytics Team**: Requires insights to support their strategic decisions and recommendations.
   - **Cyclistic Executive Team**: Needs compelling data insights and professional visualizations to approve the proposed marketing strategies.

### Deliverables

1. **Clear Statement of the Business Task**:
   - The business task is to analyze and compare the usage patterns of annual members versus casual riders to uncover actionable insights for marketing strategies.

2. **Description of Data Sources**:
   - The data sources will include historical bike trip data, which contains information on trip durations, start and end stations, bike types, and user categories (annual members vs. casual riders).

3. **Documentation of Data Cleaning or Manipulation**:
   - Data cleaning and manipulation processes will include handling missing values, removing duplicates, and transforming data into a format suitable for analysis.

4. **Summary of Analysis**:
   - A detailed summary will outline the findings from the analysis, highlighting key differences in usage patterns between annual members and casual riders.

5. **Supporting Visualizations and Key Findings**:
   - Visualizations will include graphs and charts that illustrate usage patterns, such as trip frequency, duration, and bike type preferences. Key findings will be summarized based on these visualizations.

6. **Top Three Recommendations**:
   - Based on the analysis, three actionable recommendations will be provided to help convert casual riders into annual members.
   ## Prepare

### Data Source Description

For this analysis, we will use the past 12 months of **Cyclistic’s historical trip data**, made available by **Motivate International Inc.** under an open data license. This dataset contains detailed information about bike trips, including:

- **Trip start and end times**
- **Trip durations**
- **Start and end station names**
- **Bike types** (e.g., traditional, reclining, cargo)
- **User type** (casual rider vs. annual member)

Due to privacy regulations, no personally identifiable information (PII) is included in the dataset, meaning we cannot identify riders by name, credit card information, or location of residence. This ensures compliance with data privacy laws.

### Data Location and Organization

- The data is stored in **multiple CSV files**, each representing one month of trip data. These files will need to be loaded and combined to form a comprehensive dataset covering the entire 12-month period.
- Each CSV file is structured with the same format, containing columns for trip duration, start and end stations, user types, and bike type.

### Key Data Quality Considerations

- **Licensing, Privacy, Security**: The dataset is publicly available and does not contain sensitive information, ensuring compliance with privacy and security policies.
- **Credibility**: The data is provided by a reputable source, Motivate International Inc., which manages bike-share programs in various cities. This adds to its credibility and reliability for the purpose of this analysis.
  
### Data Integrity Checks

To ensure the integrity and reliability of the dataset, the following checks will be performed:
- **Check for missing values**: Any missing or null values will be identified and handled appropriately (e.g., imputing, excluding).
- **Duplicate entries**: Duplicate rows will be removed to avoid skewing the analysis.
- **Consistency across files**: All monthly datasets will be checked for consistent column names and data types to ensure they can be properly merged.

### Data Preparation Tasks

1. **Download and Store Data**: 
   - Download the 12 monthly CSV files and store them in a designated directory for analysis.
  
2. **Data Sorting and Filtering**: 
   - Load the data into R, merge the files into one comprehensive dataset, and sort the data to ensure that it is clean and ready for analysis.
  
3. **Address Potential Bias**:
   - Since the data does not include information on where casual riders live or whether they have made multiple purchases, this may introduce bias in interpreting usage patterns.
   - The dataset focuses solely on bike trip information, so any external factors (e.g., weather, promotions) influencing bike usage are not included. This is a limitation to consider during analysis.

4. **Assess ROCCC (Reliability, Originality, Completeness, Consistency, Citedness)**:
   - **Reliability**: The data is reliable since it’s provided by an established company managing bike-share programs.
   - **Originality**: The dataset is original and specific to Cyclistic’s operations.
   - **Completeness**: The dataset spans 12 months, providing enough data to observe meaningful trends.
   - **Consistency**: Consistent formatting across months ensures smooth data merging.
   - **Citedness**: Motivate International Inc. is cited as the original source of the dataset.

### Deliverable

**Data Sources**:
- **Source**: Historical bike trip data provided by Motivate International Inc. under public license.                              
- **Format**: 12 monthly CSV files, each containing records of bike trips, user types, and related details.                        
- **Key Variables**:
   - `ride_id`, `rideable_type` (e.g., traditional bike, electric bike)
   - `started_at` and `ended_at` (trip start and end times)
   - `start_station_name` and `end_station_name` (trip station details)
   - `member_casual` (user type: casual rider or annual member)

This dataset is comprehensive for analyzing user behavior and supporting our business task of comparing annual members and casual riders.

## Prepare
##### Install and Load Needed Packages
```{r}
library(dplyr)
```
```{r}
library(tidyverse)
```
```{r}
library(skimr)
```
```{r}
library(janitor)
```
```{r}
library(lubridate)
```
```{r}
library(readr)
```
```{r}
library(ggplot2)
```
```{r}
library(tidyr)
```

## Combining and Standardizing Cyclistic Trip Data from Multiple CSV Files into a Single Data Frame

```{r}
# List of CSV file paths
csv_files <- list(
    "C:/Users/dangw/Desktop/Capstone_Project/Cyclistic_Case_Study/202408-divvy-tripdata.csv", 
    "C:/Users/dangw/Desktop/Capstone_Project/Cyclistic_Case_Study/202407-divvy-tripdata.csv", 
    "C:/Users/dangw/Desktop/Capstone_Project/Cyclistic_Case_Study/202406-divvy-tripdata.csv", 
    "C:/Users/dangw/Desktop/Capstone_Project/Cyclistic_Case_Study/202405-divvy-tripdata.csv", 
    "C:/Users/dangw/Desktop/Capstone_Project/Cyclistic_Case_Study/202404-divvy-tripdata.csv", 
    "C:/Users/dangw/Desktop/Capstone_Project/Cyclistic_Case_Study/202403-divvy-tripdata.csv", 
    "C:/Users/dangw/Desktop/Capstone_Project/Cyclistic_Case_Study/202402-divvy-tripdata.csv", 
    "C:/Users/dangw/Desktop/Capstone_Project/Cyclistic_Case_Study/202401-divvy-tripdata.csv", 
    "C:/Users/dangw/Desktop/Capstone_Project/Cyclistic_Case_Study/202312-divvy-tripdata.csv", 
    "C:/Users/dangw/Desktop/Capstone_Project/Cyclistic_Case_Study/202311-divvy-tripdata.csv", 
    "C:/Users/dangw/Desktop/Capstone_Project/Cyclistic_Case_Study/202310-divvy-tripdata.csv", 
    "C:/Users/dangw/Desktop/Capstone_Project/Cyclistic_Case_Study/202309-divvy-tripdata.csv"
)

# Function to safely read a CSV file
read_csv_safe <- function(file) {
    tryCatch({
        read.csv(file, stringsAsFactors = FALSE, na.strings = c("", "NA"))  # Set stringsAsFactors=FALSE to avoid factor issues
    }, error = function(e) {
        message(paste("Error reading file:", file))
        return(NULL)  # Return NULL in case of an error
    })
}

# Read all CSV files into a list of data frames
datasets <- lapply(csv_files, read_csv_safe)
```


```{r}
# Remove any NULL entries (from files that couldn't be read)
datasets <- datasets[!sapply(datasets, is.null)]

# Find the set of all column names across all datasets
all_columns <- Reduce(union, lapply(datasets, colnames))

# Ensure all datasets have the same columns by adding missing columns with NA
datasets <- lapply(datasets, function(df) {
    # Standardize column names to lower case and trim whitespace to avoid mismatches
    colnames(df) <- tolower(trimws(colnames(df)))
    all_columns_standardized <- tolower(trimws(all_columns))
    
    missing_cols <- setdiff(all_columns_standardized, colnames(df))
    df[missing_cols] <- NA  # Add missing columns filled with NA
    df <- df[all_columns_standardized]   # Reorder columns to match the order of 'all_columns'
    return(df)
})

# Combine all datasets by row-binding them together
yearly_data <- do.call(rbind, datasets)



```
## Process
```{r}
# View the first few rows of the resulting dataset
head(yearly_data)
```
```{r}
#Statistical summary of data
summary(yearly_data)
```

```{r}
# Check the structure of the new dataset and look for inconguencies
str(yearly_data)
```
```{r}
#List of column names
colnames(yearly_data)
```
```{r}
#number of rows in dataset?
nrow(yearly_data) 
```
```{r}
#Dimensions of the dataset?
dim(yearly_data)
```
## Cleaning


```{r}
#Remove columns that are not needed

yearly_data <- yearly_data %>%
  select(-starts_with("x"), -starts_with("start_lat"), -starts_with("start_lng"), 
         -starts_with("end_lat"), -starts_with("end_lng"))
```
```{r}
#Check the first few rows of the cleaned dataset
head(yearly_data)
```
```{r}
#Check the columns of the cleaned dataset
colnames(yearly_data)
```
```{r}
# Remove rows with any NA values
yearly_data<- drop_na(yearly_data)
```
```{r}
# Check the structure of the updated dataset
str(yearly_data)
```
## Understanding the Gaps: Acknowledging Limitations in Our Findings
Limitation in Findings: During the data pprocessing  phase, rows with missing or invalid date and time information (started_at and ended_at columns) were removed from the dataset. Although this step ensured data integrity, it may have resulted in the exclusion of valid trips that were not properly recorded in the source data. As a consequence, the analysis may not reflect the full usage patterns, particularly in cases where incomplete trip data was significant. This limitation should be considered when interpreting temporal trends, as some edge cases might have been missed. Addressing this limitation would involve ensuring that future data collection processes include consistent and accurate date-time recordings.
```{r}

# Parse 'started_at' and 'ended_at' safely and filter out invalid dates
yearly_data$started_at <- ymd_hms(yearly_data$started_at, tz = "UTC", quiet = TRUE)
yearly_data$ended_at <- ymd_hms(yearly_data$ended_at, tz = "UTC", quiet = TRUE)
yearly_data <- yearly_data %>% filter(!is.na(started_at) & !is.na(ended_at))

# Filter dates within the expected range
yearly_data <- yearly_data %>%
  filter(as.Date(started_at) >= as.Date("2023-09-01") & as.Date(started_at) <= as.Date("2024-08-31"))

# Add columns for date parts and calculate ride length
yearly_data <- yearly_data %>%
  mutate(date = as.Date(started_at),
         month = format(date, "%m"),
         day = format(date, "%d"),
         year = format(date, "%Y"),
         day_of_week = format(date, "%A"),
         ride_length = as.numeric(difftime(ended_at, started_at, units = "secs")))

# Check the result
head(yearly_data)
```
```{r}
# Calculate the 'ride_length' by finding the difference between 'ended_at' and 'started_at' timestamps.
yearly_data$ride_length <- difftime(yearly_data$ended_at, yearly_data$started_at)
str(yearly_data)
```
```{r}
# Check if the 'ride_length' column is a factor before converting it to numeric.
is.factor(yearly_data$ride_length)
```
```{r}
# Convert the 'ride_length' column from factor (or character) to numeric.
yearly_data$ride_length <- as.numeric(as.character(yearly_data$ride_length))

# Check if the 'ride_length' column is now numeric.
is.numeric(yearly_data$ride_length)

```
```{r}
# Filter out entries with ride length less than zero
yearly_data_v2 <- yearly_data %>%
  filter(ride_length >= 0)

# Check the result
head(yearly_data)

```
```{r}
# Calculate the average ride length for the cleaned dataset 'yearly_data'.
mean(yearly_data_v2$ride_length)

```
```{r}
# Calculate the median ride length for the cleaned dataset 'yearly_data'.
median(yearly_data_v2$ride_length)
```
```{r}
#longest ride
max(yearly_data_v2$ride_length)
```
```{r}
#shortest ride
min(yearly_data_v2$ride_length)
```
```{r}
# Generate a summary of the 'ride_length' column for the cleaned dataset 'yearly_data_v2'.
# The summary includes the minimum, 1st quartile, median, mean, 3rd quartile, and maximum values
summary(yearly_data_v2$ride_length)
```
```{r}
# Calculate the average ride length for each user type ('member' or 'casual') in the 'yearly_data_v2' dataset.
aggregate(yearly_data_v2$ride_length ~ yearly_data_v2$member_casual, FUN = mean)

```
```{r}
# Calculate the median ride length for each user type ('member' or 'casual') in the 'yearly_data_v2' dataset.
aggregate(yearly_data_v2$ride_length ~ yearly_data_v2$member_casual, FUN = median)

```
```{r}
# Calculate the maximum ride length for each user type ('member' or 'casual') in the 'yearly_data_v2' dataset.
aggregate(yearly_data_v2$ride_length ~ yearly_data_v2$member_casual, FUN = max)

```
```{r}
# Calculate the minimum ride length for each user type ('member' or 'casual') in the 'yearly_data_v2' dataset.
aggregate(yearly_data_v2$ride_length ~ yearly_data_v2$member_casual, FUN = min)
```
```{r}
# Calculate the average ride length for each combination of user type ('member' or 'casual') 
# and day of the week in the 'yearly_data_v2' dataset.
aggregate(yearly_data_v2$ride_length ~ yearly_data_v2$member_casual + yearly_data_v2$day_of_week, FUN = mean)
```
```{r}
# Reorder the 'day_of_week' column in 'yearly_data_v2' so that the days are in chronological order, 
# starting from Sunday through Saturday.
yearly_data_v2$day_of_week <- ordered(yearly_data_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

```
```{r}
# get the average ride time by each day for members vs casual users
aggregate(yearly_data_v2$ride_length ~ yearly_data_v2$member_casual + yearly_data_v2$day_of_week, FUN = mean)

```
```{r}
# Add a 'weekday' column indicating the day of the week when the ride started, 
# and group the data by user type ('member' or 'casual') and weekday.
yearly_data_v2 %>%  
  mutate(weekday = wday(started_at, label = TRUE)) %>%  # Create a new 'weekday' column using the 'started_at' timestamp, with day labels
  group_by(member_casual, weekday) %>%  # Group by user type and day of the week
  summarise(number_of_rides = n(),  # Count the number of rides for each group
            average_duration = mean(ride_length)) %>%  # Calculate the average ride length for each group
  arrange(member_casual, weekday)  # Arrange the results by user type and day of the week

```
# Share 
### Number of Rides by User Type and Day of the Week

```{r}
# Add a 'weekday' column, group by user type and weekday, then calculate summary statistics
yearly_data_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  # Create a new 'weekday' column from the 'started_at' timestamp with day labels
  group_by(member_casual, weekday) %>%  # Group the data by user type ('member' or 'casual') and day of the week
  summarise(number_of_rides = n(),  # Count the total number of rides for each group
            average_duration = mean(ride_length)) %>%  # Calculate the mean ride length for each group
  arrange(member_casual, weekday) %>%  # Arrange the results by user type and weekday
  
  # Create a bar chart showing the number of rides for each user type and day of the week
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge") +  # Use geom_col() to create a bar chart with side-by-side bars for each user type
  
  # Add labels and a title to the plot
  labs(title = "Number of Rides by User Type and Day of the Week",
       x = "Day of the Week", 
       y = "Number of Rides",
       fill = "User Type") +  # Label the legend as 'User Type'
  
  # Apply a minimal theme for a clean look
  theme_minimal() +
  
  # Rotate x-axis labels for better readability
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```


## Conclusion
The bar chart illustrates the **number of rides by user type** (casual vs. member) across the **days of the week**. Based on the visualized data:

- **Casual riders** tend to have a significantly higher number of rides during **weekends**, particularly on **Sundays** and **Saturdays**. This trend suggests that casual riders primarily use the service for leisure or weekend activities. The highest number of rides for casual users occurs on **Sunday**.
  
- **Member riders**, on the other hand, show a strong preference for **weekdays**, especially on **Wednesdays** and **Fridays**. This trend may indicate that members use the service more for regular commuting, possibly for work or other weekday activities.

- **Wednesday** stands out as the most popular day for **member riders**, with more than 15,000 rides, while casual riders also exhibit higher activity on this day, though less than the member users.

- **Monday** and **Thursday** are the least active days for both casual and member riders, with fewer rides recorded overall.

### Key Insights:
1. **Casual riders** are more active on weekends, likely indicating **leisure trips**.
2. **Member riders** show **consistent weekday usage**, possibly due to **commuting patterns**, particularly mid-week.

Overall, this data highlights distinct riding behaviors between casual and member users, suggesting different motivations for using the bike service.






### Average Ride Length by User Type and Day of the Week

```{r}
# Create a bar chart showing the average ride length for each user type ('member' or 'casual') by day of the week
ggplot(yearly_data_v2, aes(x = day_of_week, y = ride_length, fill = member_casual)) +
  
  # Use stat_summary to calculate and plot the mean ride length as bars, with the bars for each user type side by side
  stat_summary(fun = mean, geom = "bar", position = "dodge") +
  
  # Add labels and a title to the plot
  labs(title = "Average Ride Length by User Type and Day of the Week",
       x = "Day of the Week", 
       y = "Average Ride Length (seconds)", 
       fill = "User Type") +  # Label the legend for 'fill' with 'User Type'
  
  # Apply a minimal theme for a clean and simple look
  theme_minimal() +
  
  # Rotate the x-axis labels for better readability if the day names are long
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
## Conclusion

The bar chart illustrates the **average ride length (in seconds)** for both **casual** and **member** riders across the **days of the week**. Key findings include:

- **Casual riders** tend to have significantly longer ride durations compared to members across most days, particularly on **Tuesdays**, **Thursdays**, and **Sundays**, where the average ride length exceeds **200,000 seconds** (approximately 55 hours). This suggests that casual users may engage in more leisurely or extended rides.

- **Member riders** maintain a more consistent and shorter average ride length, usually below **100,000 seconds**. This indicates that members are likely using the service for more routine and practical trips, possibly for commuting.

- The most striking differences are observed on **Tuesdays** and **Thursdays**, where casual riders' trips are significantly longer than members'. **Sunday** also sees a high average ride duration for casual riders, further suggesting that casual users may prefer weekend leisure trips.

### Key Insights:
1. **Casual riders** exhibit much longer trip durations on average, indicating **recreational or unstructured usage patterns**.
2. **Member riders** show more **consistent and shorter trip durations**, potentially pointing to **commuting or routine travel**.
3. The difference between casual and member riders is most pronounced on **Tuesdays**, **Thursdays**, and **Sundays**, revealing distinct behavioral patterns on these days.

These insights highlight how casual and member users engage with the service differently, with casual riders favoring longer, recreational trips and members focusing on shorter, more practical rides.


### Number of Rides by Month Segmented by Rider Type

```{r}
# Create a side-by-side bar chart showing the number of rides for each user type ('member' or 'casual') by month
ggplot(yearly_data_v2, aes(x = month(started_at, label = TRUE), fill = member_casual)) +
  
  # Use geom_bar to count the number of rides for each month and display them side by side for each rider type
  geom_bar(position = "dodge", stat = "count") +
  
  # Add labels and a title to the plot
  labs(title = "Number of Rides by Month Segmented by Rider Type",
       x = "Month", 
       y = "Number of Rides",
       fill = "User Type") +  # Label the legend as 'User Type'
  
  # Apply a minimal theme for a clean look
  theme_minimal() +
  
  # Rotate x-axis labels for better readability
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```
## Conclusion

The bar chart provides an overview of the **number of rides** segmented by **casual** and **member riders** across the **months of the year**. Key observations from the chart include:

- The peak in rides occurs in the **summer and early fall months** (particularly in **July, August, and September**), where the number of rides for both casual and member users is significantly higher. This could be attributed to favorable weather conditions encouraging bike usage.

- **Member riders** tend to maintain more consistent ridership throughout the year, with noticeable spikes in **May, June, and September**. Members seem to engage in more regular usage, potentially for commuting or day-to-day activities.

- **Casual riders** experience more variability, with notable increases in ride numbers during the summer months, peaking in **September**. Casual usage drops significantly in the colder months, which suggests that casual riders may primarily use bikes for recreational or seasonal activities.

- The number of rides for **both groups** decreases significantly during the winter months (November through February), likely due to colder weather conditions that make biking less appealing.

### Key Insights:
1. **Seasonality** plays a crucial role in ride activity, with the **summer months** seeing the highest number of rides for both casual and member riders.
2. **Member riders** demonstrate more consistent usage patterns throughout the year, while **casual riders** have more seasonal variability, peaking in the summer and tapering off in the winter.
3. This trend suggests that **casual users** are more likely to use the service for **leisure** or **tourism**, while **members** might be using the service for **commuting** or **regular travel**.




```{r}
# Create a bar chart showing the number of rides for each rideable type, segmented by user type ('member' or 'casual')
ggplot(yearly_data_v2, aes(x = rideable_type, fill = member_casual)) +
  
  # Use geom_bar to count the number of rides for each rideable type
  geom_bar(position = "dodge", stat = "count") +
  
  # Add labels and a title to the plot
  labs(title = "Number of Rides by Rideable Type Segmented by User Type",
       x = "Rideable Type", 
       y = "Number of Rides",
       fill = "User Type") +  # Label the legend as 'User Type'
  
  # Apply a minimal theme for a clean look
  theme_minimal() +  # Corrected theme function
  
  # Optionally rotate x-axis labels for better readability if rideable type names are long
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
## Conclusion

The bar chart visualizes the **number of rides** segmented by **rideable type** (classic bikes vs. electric bikes) across **casual** and **member riders**. Key insights include:

- **Members** overwhelmingly prefer **classic bikes** over electric bikes, with more than **30,000 rides** on classic bikes, while electric bike usage is significantly lower in comparison.

- **Casual riders**, on the other hand, have a more balanced usage between the two types, with around **15,000 rides** on classic bikes and slightly fewer than **10,000 rides** on electric bikes.

- This trend suggests that **members** might prioritize the cost-effectiveness or availability of classic bikes for regular commuting, while **casual riders** might be more open to using electric bikes, possibly for shorter or recreational trips.

- The **electric bike adoption** by both user groups is notably higher for casual riders, indicating potential for growth in electric bike popularity, especially for non-commuting purposes.

### Key Insights:
1. **Members** show a strong preference for **classic bikes**, likely due to cost efficiency for frequent, daily rides.
2. **Casual riders** are more evenly split between **classic and electric bikes**, possibly reflecting their **preference for convenience** and **leisure**.
3. The **lower adoption** of electric bikes by **members** could be an area to explore for **future business strategies**, potentially by offering more electric bike incentives or promotions to regular users.


## Act
### Final Conclusion

The analysis of bike ride data has provided valuable insights into the usage patterns of **casual riders** versus **annual members**. Key observations include:

- **Casual riders** generally prefer longer rides and exhibit higher activity during weekends and summer months. They also show more balanced usage between classic and electric bikes, with a notable increase in electric bike usage during warmer months.

- **Annual members** tend to use the service more consistently throughout the week, particularly on weekdays. They favor classic bikes significantly more than electric bikes, indicating a preference for cost-effective and reliable options for regular commuting.

- Seasonal trends show that both casual and member riders experience a drop in ridership during the colder months, with peaks during summer.

### Recommendations

Based on these findings, the following recommendations are proposed:

1. **Increase Promotion of Electric Bikes**: To boost electric bike usage, especially among annual members, consider promotional strategies or incentives that highlight the benefits of electric bikes for both short and longer trips.

2. **Targeted Marketing Campaigns**: Develop marketing campaigns that cater to casual riders, emphasizing weekend and seasonal promotions to leverage their higher activity periods.

3. **Enhance Member Engagement**: Offer loyalty programs or additional perks to annual members to encourage year-round usage and maintain engagement during the off-peak months.

4. **Improve Seasonal Strategies**: Adjust bike availability and maintenance schedules according to seasonal trends to ensure optimal service during peak periods and to address the drop in ridership during colder months.

### Limitations

It is important to acknowledge the following limitations in this analysis:

- **Data Accuracy**: The analysis relies on the accuracy of the input data, including ride start and end times, and rideable type classifications. Any inconsistencies or inaccuracies in the dataset could affect the results.

- **Contextual Factors**: The analysis does not account for external factors such as weather conditions, special events, or changes in service availability that may influence ridership patterns.

- **Assumptions in Data Analysis**: The conclusions drawn are based on observed patterns and statistical analysis. Assumptions made during data processing, such as the exclusion of certain data points or the classification of ride types, may impact the accuracy of the insights.

- **Limited Scope**: The analysis focuses on ride length, frequency, and rideable type. Other aspects of rider behavior, such as trip purpose or demographics, were not included and could provide additional insights.

Acknowledging these limitations helps in understanding the context and constraints of the analysis, ensuring that recommendations are implemented with an awareness of potential data and analysis constraints.

## Feedback and Next Steps

As this is my first project handling a year's worth of data, there may be areas where my conclusions or findings require further refinement. In particular:

- **Data Processing**: I made certain assumptions during data cleaning, particularly around parsing date information and removing missing values. Future projects might explore more sophisticated ways of handling missing or erroneous data.
- **Ride Duration**: The calculation of ride length excluded negative or unrealistic values. While this was necessary, it might have had an impact on the overall results, potentially skewing insights.
- **Visualization Choices**: - **Visualizations**: The visualizations chosen might not fully communicate the complexity of the data. With more experience, I could select different types of graphs to present clearer or more detailed trends.


I welcome any feedback or corrections to help improve my understanding and approach for future projects.

